---
title: "GlcNac time course data analysis"
author: "Francesco Tabaro"
date: "`r Sys.Date()`"
bibliography: references.bib
link-citation: true
nocite: "*"
output:
  html_document:
    theme: cosmo
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 2
    citation_package: natbib
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r knitr_settings, include=FALSE}
knitr::knit_hooks$set(
  error = function(x, options) {
    paste('\n\n<div class="alert alert-danger">',
          gsub('##', '\n', gsub('^##\ Error', '**Error**', x)),
          '</div>', sep = '\n')
  },
  warning = function(x, options) {
    paste('\n\n<div class="alert alert-warning">',
          gsub('##', '\n', gsub('^##\ Warning:', '**Warning**', x)),
          '</div>', sep = '\n')
  },
  message = function(x, options) {
    paste('\n\n<div class="alert alert-info">',
          gsub('##', '\n', x),
          '</div>', sep = '\n')
  }
)

knitr::opts_hooks$set(
  numberLines = function(options) {
    attrs <- paste0("attr.", options$numberLines)
    options[attrs] <- lapply(options[attrs], c, ".numberLines")
    options
  }
)

knitr::opts_chunk$set(
  numberLines = c(
    "source", "output"
  )
)

knitr::opts_chunk$set(
  fig.path = 'figures/',
  fig.ext = '.png',
  fig.dev = 'png',
  fig.height = 5,
  fig.width = 5
)
```

```{r warn_user, warning=TRUE, echo=FALSE}
message("Right-click a figure and select \"open in new tab\" to get a larger view.")
```

```{r setup, message=FALSE}
library(tidyverse)
library(DESeq2)
library(patchwork)
library(ComplexUpset)
library(clusterProfiler)
library(enrichplot)
library(org.Mm.eg.db)
library(DT)
library(dendextend)
library(distances)
library(pheatmap)
library(mclust)
library(RColorBrewer)
library(decoupleR)
```

```{r define_helpers, message=FALSE, class.source='fold-hide', cache=TRUE}
get_counts <- function(file_path, normalized) {
  dds <- readRDS(file_path)
  mat <- counts(dds, normalized = normalized)
  as_tibble(mat, rownames = "gene_id")
}

get_results_table <- function(file_path){
  tp <- basename(dirname(file_path))
  
  read_csv(file_path) %>%
    dplyr::select(-1) %>%
    dplyr::mutate(time_point = tp)
  }

to_matrix <- function(df) {
  rn <- df$gene_id
  mat <- as.matrix(df[,-1])
  rownames(mat) <- rn
  mat
}

get_all_counts <- function(normalized = FALSE) list.files(
  "3t-seq/data/results/analysis/rdata/deseq2", 
  pattern = "dds.rds", 
  full.names = TRUE, 
  recursive = TRUE) %>%
  purrr::map(get_counts, normalized = normalized) %>%
  purrr::reduce(dplyr::full_join, by = "gene_id") %>%
  dplyr::select(-c(grep(".x.x.*", colnames(.)), grep(".y.*", colnames(.)))) %>%
  dplyr::rename_with(sub, pattern = ".x", replacement = "") %>%
  to_matrix

# m is a matrix with one row per gene and one column per sample.
mor <- function (m, threshold = 50, 
                 add_one = TRUE,
                 return_coefficients = FALSE, 
                 use_median = FALSE,
                 do_rounding = TRUE,
                 digits = 5
                ) {
  
  if(any(m<0, na.rm = TRUE)) stop("Negative counts detected. Aborting.", call. = FALSE)

  ## shift all cells by 1.
  if(add_one) {
    cat("Adding 1 to all cells. Turn off with add_one=FALSE.\n", file = stderr())
    m <- m + 1
  }
  
  ## build consensus sample
  if(use_median){
    consensus <- apply(m, 1, median, na.rm = TRUE)
  }else{
    consensus <- exp(rowMeans(log(m), na.rm = TRUE)) # geometric mean          
  }

  ## compute median-of-ratios  
  keep <- consensus > threshold
  mors <- apply(m, 2, function (sample) { median(sample[keep] / consensus[keep], na.rm = TRUE) })
    
  ## normalize input
  norm <- sapply(1:ncol(m), function (i) { m[,i] / mors[i] })
  
  ## round
  if (do_rounding) {
    norm <- round(norm, digits = digits)
    if(return_coefficients) mors <- round(mors, digits = digits)
  }
  
  ## keep colnames and rownames
  colnames(norm) <- colnames(m)
  rownames(norm) <- rownames(m)
  
  if(return_coefficients){
    return(list(medians = consensus, 
                coeff = mors, 
                norm = norm))
  }else{
    return(norm)    
  } 
}

create_dt <- function(x){
  require(DT)
  
  page_len <- 10
  if (nrow(x) < 10) {
    page_len <- nrow(x)
    length_menu <- "All"
  } else if (nrow(x) >= 10 & nrow(x) < 25) { 
    length_menu <- c(10)
  } else if (nrow(x) >= 25 & nrow(x) < 50) {
    length_menu <- c(10, 25)
  } else if (nrow(x) >= 50) {
    length_menu <- c(10, 25, 50)
  }
  
  dt_options <- list(pageLength = page_len)
  
  if (!is.null(length_menu)) {
    dt_options[["lengthMenu"]] <- length_menu
    dt_options[["dom"]] <- "'Blfrtip'"
  } else {
    dt_options[["dom"]] <- "'Bfrtip'"
  }
  
  dt_options[["pageLength"]] <- page_len
  
  datatable(as.data.frame(x), 
            rownames = FALSE,
            extensions = list(), 
            options = dt_options)
}

create_go_dt <- function(x) {
    if (nrow(x) > 0){
      if ("Cluster" %in% colnames(x)) {
        x <- x[,-which(colnames(x) == "Cluster")]
      }
      
      ret <- x %>% 
        as.data.frame %>%
        dplyr::select(-c(ID, pvalue, qvalue)) %>%
        dplyr::arrange(p.adjust) %>%
        dplyr::mutate(p.adjust = format(p.adjust, 
                                        scientific = TRUE, 
                                        digits = 2)) %>%
        dplyr::rename_with(tolower) %>%
        dplyr::rename(`Adjusted p-value` = p.adjust) %>%
        dplyr::rename_with(Hmisc::capitalize) %>%
        create_dt()
      return(ret)
    } else {
      warning("No GO terms enriched")
    }
}

tpm <- function(dds) {
  counts <- counts(dds, normalize = FALSE)
  lengths <- width(rowRanges(dds))
  rpk <- counts / (lengths / 1000)
  scaling_factor <- colSums(rpk) / 1e6
  tpm <- t(t(rpk) / scaling_factor)
  return(tpm)
}

get_tpm <- function(p) {
  dds <- readRDS(p)
  tpm_matrix <- tpm(dds)
  as_tibble(tpm_matrix, rownames = "gene_id")
}

get_all_tpm <- function() list.files(
  "3t-seq/data/results/analysis/rdata/deseq2", 
  pattern = "dds.rds", 
  full.names = TRUE, 
  recursive = TRUE) %>%
  purrr::map(get_tpm) %>%
  purrr::reduce(dplyr::full_join, by = "gene_id") %>%
  dplyr::select(-c(grep(".x.x.*", colnames(.)), grep(".y.*", colnames(.)))) %>%
  dplyr::rename_with(sub, pattern = ".x", replacement = "") %>%
  to_matrix

overlapGroups <- function (listInput, sort = TRUE) {
  # listInput could look like this:
  # $one
  # [1] "a" "b" "c" "e" "g" "h" "k" "l" "m"
  # $two
  # [1] "a" "b" "d" "e" "j"
  # $three
  # [1] "a" "e" "f" "g" "h" "i" "j" "l" "m"
  listInputmat    <- UpSetR::fromList(listInput) == 1
  #     one   two three
  # a  TRUE  TRUE  TRUE
  # b  TRUE  TRUE FALSE
  #...
  # condensing matrix to unique combinations elements
  listInputunique <- unique(listInputmat)
  grouplist <- list()
  # going through all unique combinations and collect elements for each in a list
  for (i in 1:nrow(listInputunique)) {
    currentRow <- listInputunique[i,]
    myelements <- which(apply(listInputmat,1,function(x) all(x == currentRow)))
    attr(myelements, "groups") <- currentRow
    grouplist[[paste(colnames(listInputunique)[currentRow], collapse = ":")]] <- myelements
    myelements
    # attr(,"groups")
    #   one   two three 
    # FALSE FALSE  TRUE 
    #  f  i 
    # 12 13 
  }
  if (sort) {
    grouplist <- grouplist[order(sapply(grouplist, function(x) length(x)), decreasing = TRUE)]
  }
  attr(grouplist, "elements") <- unique(unlist(listInput))
  return(grouplist)
  # save element list to facilitate access using an index in case rownames are not named
}
```

We will begin by importing data and performing median of ratios normalization of the count matrix. The table below shows the initial number of genes and samples.

```{r import_data, cache=TRUE}
all_counts <- get_all_counts(normalized = FALSE)
all_counts <- all_counts[apply(all_counts, 1, function(x) all(is.finite(x))),]
all_counts_mor <- mor(all_counts, return_coefficients = FALSE, do_rounding = FALSE)
all_tpm <- get_all_tpm()

data.frame(row.names = c("genes", "samples"), values = dim(all_counts))
```

We will also import all DESeq2 results table as computed from 3t-seq.

```{r import_results_table, cache=TRUE, message=FALSE}
all_deseq2_results_tables <- list.files(
  "3t-seq/data/results/analysis/tables/deseq2", 
  pattern = "results.shrink.csv", 
  full.names = TRUE, 
  recursive = TRUE) %>%
  purrr::map(get_results_table) %>%
  purrr::reduce(bind_rows) %>%
  dplyr::mutate(time_point = factor(time_point, 
                                    levels = c("0_30", "0_2",
                                               "0_4", "0_8",
                                               "0_24", "0_48",
                                               "0_7D", "0_WT",
                                               "WT_WT7D")),
                log2FoldChange = ifelse(time_point == "0_WT", -log2FoldChange, log2FoldChange))

as.data.frame(all_deseq2_results_tables)
```

# Combine pairwise results

We will produce a compound plot merging all Volcano and MA plots for pairwise analysis performed by 3t-seq.

We begin by setting log2 fold-change and adjusted p-values threshold to call differentially expressed genes (DEG).

```{r set_lfc_pval_thres, echo=FALSE, cache=TRUE}
lfc_thr <- 0
pval_thr <- 1e-2

data.frame(row.names = c("log2FoldChange threshold", "Adjusted p-value threshold"), values = c(lfc_thr, pval_thr))
```

With these thresholds we proceed to call differentially expressed genes (DEG) in each time point. We produce a heatmap to summarize gene expression changes across all time points.

```{r call_deg, cache=TRUE, fig.height=8, fig.width=8}
all_deg <- all_deseq2_results_tables %>%
  dplyr::mutate(deg = abs(log2FoldChange) > lfc_thr 
                & ifelse(is.na(padj), pvalue < pval_thr, padj < pval_thr),
                direction = sign(log2FoldChange))

deg_mat <- all_counts_mor[rownames(all_counts_mor) %in% (all_deg %>% 
                            dplyr::pull(gene_id) %>% 
                            unique()),
                          !grepl("WT", colnames(all_counts_mor))]

colorder <- rep(c("0h", "30min", "2h", "4h", "8h", "24h", "48h", "7D"), each = 3)
replicates <- paste0(rep("rep", 3), 1:3)
labs <- expand.grid(colorder, replicates)
labs <- paste(labs$Var1, labs$Var2, sep="_")

hcc <- hclust(
  distances::distance_matrix(
    distances::distances(
      scale(t(deg_mat), 
            center = rowMeans(deg_mat), 
            scale = rowSds(deg_mat)), 
      id_variable = colnames(deg_mat)
    )),
  method = "ward.D")
hcc <- dendextend::rotate(hcc, order = labs)  
  
pheatmap::pheatmap(deg_mat,
                   scale = "row",
                   cluster_cols = hcc,
                   show_rownames = FALSE)
```

## Volcanos and MA plots

We produce a side-by-side visualization of gene expression through Volcano plot and MA plots.

```{r 3t-seq_merge_volcano_ma_plot, fig.width=20, fig.height=10, cache=TRUE, message=FALSE, warning=FALSE}


pick_color <- function (deg_status, direction) {
  if (length(deg_status) != length(direction)){ 
    stop("Input vectors must have the same length.")
  }
  color <- rep("grey", length(deg_status))
  color[deg_status & direction == 1] <- "blue"
  color[deg_status & direction == -1] <- "red"
  return(color)
}

labels <- c(`0_30` = "30min vs 0h",
            `0_2` = "2h vs 0h",
            `0_4` = "4h vs 0h",
            `0_8` = "8h vs 0h",
            `0_24` = "24h vs 0h",
            `0_48` = "48h vs 0h",
            `0_7D` = "7 days vs 0h",
            `0_WT` = "0h vs WT",
            `WT_WT7D` = "WT 7 days vs WT")

volcano <- all_deg %>%
  mutate(color = pick_color(deg, direction)) %>%
  ggplot(aes(
    x = log2FoldChange, 
    y = -log10(padj), 
    color = color)
    ) +
  geom_point(size = .75) +
  geom_hline(
    yintercept = -log10(pval_thr), 
    color = "black") +
  geom_vline(
    xintercept = c(-lfc_thr, lfc_thr), 
    color = "black") +
  scale_color_identity() +
  xlab("log2 Fold Change") +
  ylab("-log10(adjusted p-value)") +
  facet_wrap(facets = vars(time_point), ncol = 3,
             labeller = labeller(time_point = labels)) +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1)

ma_plot <- all_deg %>%
  mutate(color = pick_color(deg, direction)) %>%
  ggplot(aes(x = log10(baseMean), y = log2FoldChange, color = color)) +
  geom_point(size = .75) +
  geom_hline(
    yintercept = c(-lfc_thr, lfc_thr), 
    color = "black") +
  scale_color_identity() +
  xlab("Mean expression (log10)") +
  ylab("log2 fold change") +
  facet_wrap(facets = vars(time_point), ncol = 3,
             labeller = labeller(time_point = labels)) + 
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1)

volcano + ma_plot
```

## Plot number of DEG across time points

```{r number_of_deg_per_comparison, fig.height=10, fig.width=10, cache=TRUE}
all_deg %>%
  dplyr::group_by(time_point) %>%
  dplyr::summarize(
    upregulated = sum(deg & direction == 1, na.rm=TRUE),
    downregulated = sum(deg & direction == -1, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(cols = c(upregulated, downregulated)) %>%
  ggplot2::ggplot(aes(x = time_point, 
                      y = value, 
                      fill = name, 
                      label = value)) +
  ggplot2::geom_bar(stat = "identity", 
                    position = "dodge", 
                    color = "white") +
  ggplot2::geom_text(mapping = aes(y = value + 50), 
                     position = position_dodge(width = 0.9)) +
  ggplot2::scale_x_discrete(labels = c(
    `0_30` = "0h vs 30min",
    `0_2` = "0h vs 2h",
    `0_4` = "0h vs 4h",
    `0_8` = "0h vs 8h",
    `0_24` = "0h vs 24h",
    `0_48` = "0h vs 48h",
    `0_7D` = "0h vs 7 days",
    `0_WT` = "0h vs WT 0h",
    `WT_WT7D` = "WT 0h vs WT 7 days")) +
  ggplot2::scale_fill_manual(name = "Direction", 
                             labels = Hmisc::capitalize, 
                             values = c(downregulated = "red", 
                                        upregulated = "blue")) +
  ggplot2::xlab("Comparison") +
  ggplot2::ylab("Number of genes") +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(aspect.ratio = 1, 
                 axis.text.x = element_text(angle = 45, hjust = 1),
                 legend.position = "bottom")
```

## Compute DEG shared across time points

```{r upset_constants, cache=TRUE, class.source='fold-hide', warning=FALSE}
my_theme <- upset_modify_themes(
  list(
    'default' = theme(
      text = element_text(size = 16),
      axis.text.x = element_text(size = 16)),
    'intersections_matrix' = theme(
      text = element_text(size = 16)
    ),
    'Intersection size' = theme(
      text = element_text(size = 16)
    ),
    'overall_sizes' = theme(
      text = element_text(size = 16),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  ))

my_intersection_matrix <- intersection_matrix(
            geom = geom_point(shape = 19, size = 2.5),
            segment = geom_segment(
              linetype = 'dotted', size = 0.25, color = "black"),
            outline_color = list(active = 'black', inactive = 'grey90'))
```

To inspect how DEG are shared among time points and controls we will produce a [UpSet plot](https://en.wikipedia.org/wiki/UpSet_plot).

We will produce two separate visualizations, one for upregulated, the other for downregulated genes.

We will remove `0_7D` contrast as it generates a lot of noise and is not relevant.

```{r upset_init, cache=TRUE}
dat <- all_deg %>%
  dplyr::filter(deg
                & ! time_point == "0_7D") %>%
  dplyr::mutate(direction = sign(log2FoldChange)) %>%
  dplyr::select(gene_id, direction, time_point) %>%
  dplyr::group_by(direction)
  
keys <- dat %>%
  dplyr::group_data() %>%
  dplyr::pull(direction)

venn.data <- dat %>%
  dplyr::group_by(direction) %>%
  dplyr::group_split() %>%
  purrr::set_names(keys)

vd_complexUpset <- venn.data %>%
  purrr::map(~.x %>% 
               dplyr::select(-direction) %>% 
               dplyr::mutate(val = TRUE) %>%
               tidyr::pivot_wider(
                 id_cols = gene_id, 
                 names_from = time_point, 
                 values_from = val, 
                 values_fill = FALSE))
```

### Upregulated

We will produce a UpSet plot showing all intersections with at least 5 elements.

```{r upset_all_upregulated, fig.height=8, fig.width=14, cache=TRUE}
venn.upregulated <- vd_complexUpset[["1"]] %>%
  dplyr::select(-gene_id) %>%
  as.data.frame

time_points <- c("0_30", "0_2", "0_4", "0_8", "0_24", "0_48", "0_WT", "WT_WT7D")

upset(
  venn.upregulated,
  intersect = rev(time_points),
  sort_sets = FALSE,
  name = NULL,
  width_ratio = 0.1,
  sort_intersections_by = c("degree", "cardinality"),
  set_sizes = upset_set_size(),
  wrap = FALSE,
  themes = my_theme,
  min_size = 5,
  matrix = my_intersection_matrix
  )
```

We will focus on a subset of more interesting intersections, ie. intersctions of consecutive time points.

```{r upset_selected_upregulated, fig.height=8, fig.width=14, cache=TRUE}

grouped_data <- venn.data[["1"]] %>% 
  dplyr::group_by(time_point)

keys <- grouped_data %>%
  dplyr::group_keys() %>%
  dplyr::pull("time_point")
keys <- paste0("X", keys)

L <- grouped_data %>% 
  dplyr::group_split() %>% 
  purrr::set_names(keys) %>%
  purrr::map(dplyr::pull, "gene_id") 

UpSetR::upset(
  UpSetR::fromList(L),
  intersections = list(
    list(`0_30` = "X0_30", `0_2` = "X0_2", `0_4` = "X0_4", `0_8` = "X0_8", `0_24` = "X0_24", `0_48` = "X0_48", `0_WT` = "X0_WT"),
    list(`0_30` = "X0_30", `0_2` = "X0_2", `0_4` = "X0_4", `0_8` = "X0_8", `0_24` = "X0_24", `0_48` = "X0_48"),
    list(`0_2` = "X0_2", `0_4` = "X0_4", `0_8` = "X0_8", `0_24` = "X0_24", `0_48` = "X0_48"),
    list(`0_4` = "X0_4", `0_8` = "X0_8", `0_24` = "X0_24", `0_48` = "X0_48"),
    list(`0_8` = "X0_8", `0_24` = "X0_24", `0_48` = "X0_48"),
    list(`0_24` = "X0_24", `0_48` = "X0_48")
  ),
  sets = c("X0_30","X0_2","X0_4","X0_8","X0_24","X0_48", "X0_WT"),
  keep.order = TRUE,
  order.by = "degree",
  decreasing = TRUE,
  text.scale = 1.5
  )
```

Next, we will pull out the gene lists corresponding to these intersections and compute GO terms enrichments on those.

```{r go_enrichment_upregulated_shared_genes, cache=TRUE, fig.height=9, fig.width=7, cache=TRUE}

x <- overlapGroups(L %>% purrr::discard_at("XWT_WT7D"))

genes <- list(
  `X0_30:X0_2:X0_4:X0_8:X0_24:X0_48:X0_WT` = attr(x, "elements")[x[["X0_30:X0_2:X0_4:X0_8:X0_24:X0_48:X0_WT"]]],
  `X0_30:X0_2:X0_4:X0_8:X0_24:X0_48` = attr(x, "elements")[x[["X0_30:X0_2:X0_4:X0_8:X0_24:X0_48"]]],
  `X0_2:X0_4:X0_8:X0_24:X0_48` = attr(x, "elements")[x[["X0_2:X0_4:X0_8:X0_24:X0_48"]]],
  `X0_4:X0_8:X0_24:X0_48` = attr(x, "elements")[x[["X0_4:X0_8:X0_24:X0_48"]]],
  `X0_8:X0_24:X0_48` = attr(x, "elements")[x[["X0_8:X0_24:X0_48"]]],
  `X0_24:X0_48` = attr(x, "elements")[x[["X0_24:X0_48"]]]
) %>%
  purrr::map2(.y = names(.), ~tibble(gene_id = .x, intersection = .y)) %>%
  purrr::reduce(bind_rows) %>%
  dplyr::mutate(gene_id = paste0("MGI:", gene_id))

universe <- all_deg %>%
  pull(gene_id) %>%
  unique() %>%
  paste0("MGI:", .)

ck_upset_up <- compareCluster(gene_id ~ intersection, 
                     data = genes, 
                     fun = enrichGO,
                     OrgDb = 'org.Mm.eg.db',
                     keyType = "MGI",
                     ont = "BP",
                     universe = universe,
                     minGSSize = 15)

ck_upset_up <- setReadable(ck_upset_up, OrgDb = org.Mm.eg.db, keyType = "MGI")

dotplot(ck_upset_up, 
        x = "intersection", 
        showCategory = 2) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size=16),
        axis.title = element_text(size=16),
        axis.text.y = element_text(size=14))
```

Tables below show detailed results.

```{r split_ck_shared_degs_up, echo=FALSE, cache=TRUE}
splitted_upset_up <- as.data.frame(ck_upset_up) %>%
  group_by(Cluster) %>%
  group_split()
```

###  {.tabset}

#### 0_30 & 0_2 & 0_4 & 0_8 & 0_24 & 0_48 & 0_WT

```{r go_up_X0_30:X0_2:X0_4:X0_8:X0_24:X0_48:X0_WT, cache=TRUE, echo=FALSE}
create_go_dt(splitted_upset_up[[3]])
```

#### 0_30 & 0_2 & 0_4 & 0_8 & 0_24 & 0_48

```{r go_up_X0_30:X0_2:X0_4:X0_8:X0_24:X0_48, cache=TRUE, echo=FALSE}
create_go_dt(splitted_upset_up[[2]])
```

#### 0_2 & 0_4 & 0_8 & 0_24 & 0_48

```{r go_up_X0_2:X0_4:X0_8:X0_24:X0_48, cache=TRUE, echo=FALSE}
warning("No enriched terms.")
```

#### 0_4 & 0_8 & 0_24 & 0_48

```{r go_up_X0_4:X0_8:X0_24:X0_48, cache=TRUE, echo=FALSE}
warning("No enriched terms.")
```

#### 0_8 & 0_24 & 0_48

```{r go_up_X0_8:X0_24:X0_48, cache=TRUE}
create_go_dt(splitted_upset_up[[4]])
```

#### 0_24 & 0_48

```{r go_up_X0_24:X0_48, cache=TRUE}
create_go_dt(splitted_upset_up[[1]])
```

###  {.unnumbered}

### Downregulated

```{r upset_all_downregulated, fig.height=8, fig.width=14, cache=TRUE}

venn.downregulated <- vd_complexUpset[["-1"]] %>%
  dplyr::select(-gene_id)

time_points <- colnames(venn.downregulated)

upset(
  venn.downregulated,
  intersect = rev(time_points),
  sort_sets = FALSE,
  name = NULL,
  width_ratio = 0.1,
  sort_intersections_by = c("degree", "cardinality"),
  set_sizes = upset_set_size(),
  wrap = FALSE,
  themes = my_theme,
  min_size = 5,
  matrix = my_intersection_matrix
  )
```

Similarly to upregulated genes, we will produce a UpSet plot showing the number of DEG shared across consecutive time points.

```{r upset_selected_downregulated, fig.height=8, fig.width=14, cache=TRUE}

grouped_data <- venn.data[["-1"]] %>% 
  dplyr::group_by(time_point)

keys <- grouped_data %>%
  dplyr::group_keys() %>%
  dplyr::pull("time_point")
keys <- paste0("X", keys)

L <- grouped_data %>% 
  dplyr::group_split() %>% 
  purrr::set_names(keys) %>%
  purrr::map(dplyr::pull, "gene_id") 

UpSetR::upset(
  UpSetR::fromList(L),
  intersections = list(
    list(`0_30` = "X0_30", `0_2` = "X0_2", `0_4` = "X0_4", 
         `0_8` = "X0_8", `0_24` = "X0_24", `0_48` = "X0_48"),
    list(`0_2` = "X0_2", `0_4` = "X0_4", `0_8` = "X0_8", 
         `0_24` = "X0_24", `0_48` = "X0_48"),
    list(`0_4` = "X0_4", `0_8` = "X0_8", `0_24` = "X0_24", 
         `0_48` = "X0_48"),
    list(`0_8` = "X0_8", `0_24` = "X0_24", `0_48` = "X0_48"),
    list(`0_24` = "X0_24", `0_48` = "X0_48")
  ),
  sets = c("X0_30","X0_2","X0_4","X0_8","X0_24","X0_48"),
  keep.order = TRUE,
  order.by = "freq",
  decreasing = FALSE,
  text.scale = 1.5
  )
```

We will compute GO Enrichment also for the sets of shared downregulated genes.

```{r go_enrichment_downregulated_shared_genes, fig.height=9, fig.width=7, cache=TRUE}

x <- overlapGroups(L %>% purrr::discard_at("XWT_WT7D"))

genes <- list(
  `X0_30:X0_2:X0_4:X0_8:X0_24:X0_48:X0_WT` = attr(x, "elements")[x[["X0_30:X0_2:X0_4:X0_8:X0_24:X0_48:X0_WT"]]],
  `X0_30:X0_2:X0_4:X0_8:X0_24:X0_48` = attr(x, "elements")[x[["X0_30:X0_2:X0_4:X0_8:X0_24:X0_48"]]],
  `X0_2:X0_4:X0_8:X0_24:X0_48` = attr(x, "elements")[x[["X0_2:X0_4:X0_8:X0_24:X0_48"]]],
  `X0_4:X0_8:X0_24:X0_48` = attr(x, "elements")[x[["X0_4:X0_8:X0_24:X0_48"]]],
  `X0_8:X0_24:X0_48` = attr(x, "elements")[x[["X0_8:X0_24:X0_48"]]],
  `X0_24:X0_48` = attr(x, "elements")[x[["X0_24:X0_48"]]]
) %>%
  purrr::map2(.y = names(.), ~tibble(gene_id = .x, intersection = .y)) %>%
  purrr::reduce(bind_rows) %>%
  dplyr::mutate(gene_id = paste0("MGI:", gene_id))

universe <- all_deg %>%
  pull(gene_id) %>%
  unique() %>%
  paste0("MGI:", .)

ck_upset_down <- compareCluster(gene_id ~ intersection, 
                     data = genes, 
                     fun = enrichGO,
                     OrgDb = 'org.Mm.eg.db',
                     keyType = "MGI",
                     ont = "BP",
                     universe = universe,
                     minGSSize = 15)

ck_upset_down <- setReadable(ck_upset_down, OrgDb = org.Mm.eg.db, keyType = "MGI")

dotplot(ck_upset_down, 
        x = "intersection", 
        showCategory = 2) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size=16),
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14))
```

Tables below show detailed results.

```{r split_ck_shared_degs_down, echo=FALSE, cache=TRUE}
splitted_upset_down <- as.data.frame(ck_upset_down) %>%
  group_by(Cluster) %>%
  group_split()
```

###  {.tabset}

#### 0_30 & 0_2 & 0_4 & 0_8 & 0_24 & 0_48 & 0_WT

```{r go_down_X0_30:X0_2:X0_4:X0_8:X0_24:X0_48:X0_WT, cache=TRUE, echo=FALSE}
create_go_dt(splitted_upset_down[[4]])
```

#### 0_30 & 0_2 & 0_4 & 0_8 & 0_24 & 0_48

```{r go_down_X0_30:X0_2:X0_4:X0_8:X0_24:X0_48, cache=TRUE, echo=FALSE}
create_go_dt(splitted_upset_down[[3]])
```

#### 0_2 & 0_4 & 0_8 & 0_24 & 0_48

```{r go_down_X0_2:X0_4:X0_8:X0_24:X0_48, cache=TRUE, echo=FALSE}
create_go_dt(splitted_upset_down[[1]])
```

#### 0_4 & 0_8 & 0_24 & 0_48

```{r go_down_X0_4:X0_8:X0_24:X0_48, cache=TRUE, echo=FALSE}
create_go_dt(splitted_upset_down[[5]])
```

#### 0_8 & 0_24 & 0_48

```{r go_down_X0_8:X0_24:X0_48, cache=TRUE, echo=FALSE}
warning("No enriched terms.")
```

#### 0_24 & 0_48

```{r go_down_X0_24:X0_48, cache=TRUE, echo=FALSE}
create_go_dt(splitted_upset_down[[2]])
```

###  {.unnumbered}

## Plot OGT and OGA genes expression over time

```{r plot_oga_ogt_expression_over_time, message=FALSE, fig.width=10, fig.height=10, cache=TRUE}

ogt <- "MGI:1339639"
oga <- "MGI:1932139"

labs <- c("MGI:1339639" = "Ogt", 
          "MGI:1932139" = "Oga")

all_tpm %>%
  dplyr::as_tibble(rownames = "gene_id") %>%
  dplyr::filter(gene_id %in% c(ogt, oga)) %>%
  tidyr::pivot_longer(cols = -gene_id) %>%
  dplyr::mutate(
    tp = sapply(str_split(name, "_"), `[`, 1),
    rep = sapply(str_split(name, "_"), `[`, 2)
  ) %>%
  dplyr::filter(!grepl("WT", tp)) %>%
  ggplot2::ggplot(mapping = aes(x = tp, 
                               y = value, 
                               group = gene_id, 
                               color = gene_id, 
                               fill = gene_id)) +
  ggplot2::geom_jitter(width = 0.1) +
  ggplot2::geom_smooth(method = stats::loess) + 
  ggplot2::scale_color_discrete(labels = labs, name = "Gene") +
  ggplot2::scale_fill_discrete(labels = labs, name = "Gene") + 
  ggplot2::scale_x_discrete(name = "Time points",
                           labels = c("0h", "30min", 
                                      "2h", "4h", 
                                      "8h", "24h", 
                                      "48h", "7D")) +
  ggplot2::ylab("TPM") +
  ggplot2::theme_bw(base_size = 16) +
  ggplot2::theme(aspect.ratio = 9/16,
                 legend.position = "bottom")
```

# Plot RPL and RPS genes expression over time

We manually searched for ribosomal genes on the Mouse Genetics Informatics (MGI) website. We downloaded the results in CSV format and will use it to subset our gene expression table for ribosomal genes of interest.

```{r get_MGI, message=FALSE, cache=TRUE}
mgi_search_results <- read_csv("MGI_Features_20240610_071948.csv") %>%
  dplyr::filter(Type != "pseudogene" 
                & Start != "syntenic" 
                & !`Best Match Type` %in% c("Protein Domain", "old name", "Process") 
                & Chr != "MT")
```

```{r datatable_RPL_RPS_genes, echo=FALSE, cache=TRUE}
create_dt(mgi_search_results)
```

We will produce a clustered heatmap of Z-scores to show expression trends.

```{r heatmap_RPL_RPS_genes, fig.width=8, fig.height=8, cache=TRUE}
selected_mgi <- mgi_search_results %>% pull("MGI ID")

print(sum(selected_mgi %in% rownames(all_counts_mor)) / length(selected_mgi) * 100)

print(setdiff(selected_mgi, rownames(all_counts_mor)))

mat <- all_counts_mor[rownames(all_counts_mor) %in% selected_mgi,
                      !grepl("WT", colnames(all_counts_mor))] 

colorder <- rep(c("0h", "30min", "2h", "4h", "8h", "24h", "48h", "7D"), each = 3)
replicates <- paste0(rep("rep", 3), 1:3)
labs <- expand.grid(colorder, replicates)
labs <- paste(labs$Var1, labs$Var2, sep="_")

hcc <- hclust(
  distances::distance_matrix(
    distances::distances(
      scale(t(mat), 
            center = rowMeans(mat), 
            scale = rowSds(mat)), 
      id_variable = colnames(mat)
    )),
  method = "ward.D")
hcc <- dendextend::rotate(hcc, order = labs)  

hcr <- hclust(
  d = dist(t(scale(t(mat), center = rowMeans(mat), scale= rowSds(mat))), method = "euclidean"),
  method = "complete"
)

clusters <- data.frame(row.names = rownames(mat),
                       gene_id = rownames(mat), 
                       cluster = as.factor(cutree(hcr, 2)))

pheatmap::pheatmap(
  mat,
  cluster_cols = hcc,
  cluster_rows = hcr,
  scale = "row",
  show_rownames = FALSE,
  main = "RPL and RPS genes (Z-score)",
  cutree_rows = 2,
  annotation_row = subset(clusters, select = -1)
)
```

```{r plot_rpl_rps_by_cluster, fig.width=10, fig.height=5, cache=TRUE, message=FALSE}
mat <- t(scale(t(mat), center = rowMeans(mat), scale = rowSds(mat)))

dat <- mat %>%
  dplyr::as_tibble(rownames = "gene_id") %>%
  dplyr::inner_join(clusters) %>%
  tidyr::pivot_longer(cols = -c(gene_id, cluster)) %>%
  dplyr::mutate(time_point = sapply(str_split(name, "_"), `[`, 1),
         replicate = sapply(str_split(name, "_"), `[`, 2),
         time_point = factor(time_point, levels = c("0h", "30min", "2h", "4h", "8h", "24h", "48h", "7D"))) %>%
  dplyr::group_by(gene_id, time_point, cluster) %>%
  dplyr::summarise(exp = mean(value)) 

avg <- dat %>%
  group_by(time_point, cluster) %>%
  summarise(mean_exp = median(exp))

dat %>%
  ggplot2::ggplot(aes(time_point, exp)) + 
  ggplot2::geom_smooth(data = avg, mapping = aes(x =  time_point, y = mean_exp, group = cluster), color = "red") +
  ggplot2::geom_boxplot(fill = NA) +
  ggplot2::facet_wrap(~ cluster, labeller = labeller(cluster = c(`1` = "Cluster 1", `2` = "Cluster 2"))) +
  ggplot2::ylab("Expression Z-score") +
  ggplot2::theme_bw(base_size = 16) + 
  ggplot2::theme(aspect.ratio = 1)
```

Tables below show the MGI search results table split by cluster.

###  {.tabset}

#### Cluster 1

```{r print_rpl_rps_cluster_1, echo=FALSE, cache=TRUE}
cluster1 <- clusters %>% filter(cluster == 1) %>% pull(gene_id)

create_dt(mgi_search_results %>%
  filter(`MGI ID` %in% cluster1))
```

#### Cluster 2

```{r print_rpl_rps_cluster_2, echo=FALSE, cache=TRUE}
cluster2 <- clusters %>% filter(cluster == 2) %>% pull(gene_id)

create_dt(mgi_search_results %>%
  filter(`MGI ID` %in% cluster2))
```

###  {.unnumbered}

We will also produce a Volcano and MA plots highlighting significantly deregulated gene coding for ribosomal proteins.

```{r ma_volcano_RPL_RPS_genes, fig.width=20, fig.height=10, warning=FALSE, cache=TRUE}
pick_color <- function (gene_id, deg_status, direction) {
  if (length(deg_status) != length(direction)){ 
    stop("Input vectors must have the same length.")
  }
  color <- rep("grey", length(deg_status))
  color[deg_status & direction == 1] <- "dodgerblue"
  color[deg_status & direction == -1] <- "salmon"
  color[deg_status
        & direction == 1
        & gene_id %in% mgi_search_results$`MGI ID`] <- "blue"
  color[deg_status
        & direction == -1
        & gene_id %in% mgi_search_results$`MGI ID`] <- "red"
  color[!deg_status
        & direction == 1
        & gene_id %in% mgi_search_results$`MGI ID`] <- "darkblue"
  color[!deg_status
        & direction == -1
        & gene_id %in% mgi_search_results$`MGI ID`] <- "darkred"
  return(color)
}

point_size <- 0.5

legend_labels <-  c(dodgerblue = "Upregulated",
                    salmon = "Downregulated",
                    grey = "Gene",
                    darkred = "Ribosomal upregulated N.S.",
                    darkblue = "Ribosomal downregulated N.S.",
                    red = "Ribosomal upregulated",
                    blue = "Ribosomal downregulated")

labels <- c(`0_30` = "30min vs 0h",
            `0_2` = "2h vs 0h",
            `0_4` = "4h vs 0h",
            `0_8` = "8h vs 0h",
            `0_24` = "24h vs 0h",
            `0_48` = "48h vs 0h",
            `0_7D` = "7 days vs 0h",
            `0_WT` = "0h vs WT",
            `WT_WT7D` = "WT 7 days vs WT")


volcano <- all_deg %>%
  mutate(color = pick_color(gene_id, deg, direction)) %>%
  ggplot(aes(
    x = log2FoldChange, 
    y = -log10(padj), 
    color = color)
    ) +
  geom_point(size = point_size) +
  geom_point(
    data = all_deg %>%
      mutate(color = pick_color(gene_id, deg, direction)) %>%
      filter(gene_id %in% mgi_search_results$`MGI ID`),
    size = point_size
  ) +
  geom_hline(
    yintercept = -log10(pval_thr), 
    color = "black") +
  geom_vline(
    xintercept = c(-lfc_thr, lfc_thr), 
    color = "black") +
  scale_color_identity(
    guide = "legend", 
    name = "", 
    labels = legend_labels) +
  guides(colour = guide_legend(override.aes = list(size = 4))) + 
  xlab("log2 Fold Change") +
  ylab("-log10(adjusted p-value)") +
  facet_wrap(facets = vars(time_point), ncol = 3,
             labeller = labeller(time_point = labels)) + 
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1, 
        legend.position = "bottom")

ma_plot <- all_deg %>%
  mutate(color = pick_color(gene_id, deg, direction)) %>%
  ggplot(aes(x = log10(baseMean), y = log2FoldChange, color = color)) +
  geom_point(size = point_size) +
  geom_point(
    data = all_deg %>%
      mutate(color = pick_color(gene_id, deg, direction)) %>%
      filter(gene_id %in% mgi_search_results$`MGI ID`),
    size = point_size
  ) +
  geom_hline(
    yintercept = c(-lfc_thr, lfc_thr), 
    color = "black") +
  scale_color_identity(
    guide = "legend", 
    name = "", 
    labels = legend_labels) +
  guides(colour = guide_legend(override.aes = list(size = 4))) + 
  xlab("Mean expression (log10)") +
  ylab("log2 fold change") +
  facet_wrap(facets = vars(time_point), ncol = 3,
             labeller = labeller(time_point = labels)) + 
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1, 
        legend.position = "bottom")

volcano + ma_plot + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

# Principal Component Analysis

## All samples

We will compute PCA using all replicates from all samples across all time points. We will compute it on median-of-ratios normalized counts.

The plot shows a large contribution of time on PC1 and the genetic background emerge on PC2.

```{r PCA, fig.width=7, fig.height=5, cache=TRUE}
do_pca <- function(mat, legend_labels) {
  pc <- prcomp(t(mat), center = TRUE, scale = TRUE)
  
  tp <- sapply(strsplit(colnames(mat), "_", fixed = TRUE), `[[`, 1)
  tp <- factor(tp, levels = legend_labels)
  colors <- as.numeric(tp)
  
  is_wt <- grepl("WT", colnames(mat))
  shape <- ifelse(is_wt, 17, 19)
  
  perc_sdev <- pc$sdev / sum(pc$sdev) * 100
  pc1 <- sprintf("PC1 (%.1f%%)", perc_sdev[1])
  pc2 <- sprintf("PC2 (%.1f%%)", perc_sdev[2])
  
  par(mar = c(4,4,2,8.5), pty = "s")
  plot(pc$x[,1], pc$x[,2], col = colors, pch = shape, xlab = pc1, ylab = pc2)
  lx <- ceiling(max(pc$x[,1])) + ceiling(max(pc$x[,1]) * 0.1)
  ly <- ceiling(max(pc$x[,2]))
  legend(x = lx, y = ly,
    legend = levels(tp), 
    col = 1:length(levels(tp)), 
    pch = ifelse(grepl("WT", levels(tp)), 17, 19),
    horiz = FALSE, 
    ncol = 2, 
    xpd = TRUE)
}

do_pca(all_counts_mor, 
       legend_labels = c("0h", "30min", "2h", "4h", "8h", "24h", "48h", "7D", "WT", "WT7D"))
```

## Without WT

We will explore how the median of ratios normalization affects PCA. We will go back to the raw counts, remove WT samples, re-run the median of ratios normalization and PCA.

The scatterplot of the first two components looks substantially different than previous iterations. The effect of time is spread in diagonal between PC1 and PC2.

```{r PCA_no_wt_with_mor, fig.width=7.25, fig.height=5, cache=TRUE}
mat <- all_counts
mat <- mat[,-grep("WT", colnames(mat))]
mat_mor <- mor(mat, return_coefficients = FALSE, do_rounding = FALSE)

do_pca(mat, 
       legend_labels = c("0h", "30min", "2h", "4h", "8h", "24h", "48h", "7D"))
```

# Time-course analysis

We will start by median-of-ratios (mor) normalized counts and compute Z-scores. We will filter for genes with at least 50 mor-normalized read counts and with a variance value above the lower quartile value of the global variances distribution. We will compute Z-scores on this filtered gene expression matrix.

```{r time_course_analysis, cache=TRUE, message=FALSE}
colorder <- c("0h", "30min", "2h", "4h", "8h", "24h", "48h")

mat <- all_counts_mor[,-c(grep("7D", colnames(all_counts_mor)), 
                          grep("WT", colnames(all_counts_mor)))]
mat <- mat[rowSums(mat) > 50,]

mat <- mat[rowVars(mat) > quantile(rowVars(mat), .25),]

# t0_median <- rowMedians(mat[,grep("0h", colnames(mat))])

norm <- t(scale(t(mat),
        center = rowMeans(mat),
        scale = rowSds(mat))) 

as_matrix <- function(tbl) {
  rn <- tbl$gene_id
  mat <- as.matrix(tbl[,-which(colnames(tbl) == "gene_id")])
  rownames(mat) <- rn
  mat <- mat[,colorder]
  return(mat)
}

norm <- norm %>%
  as_tibble(rownames = "gene_id") %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "sample",
    values_to = "expression"
  ) %>%
    mutate(time_point = sapply(str_split(sample, "_"), `[`, 1),
         replicate = sapply(str_split(sample, "_"), `[`, 2)) %>%
  group_by(gene_id, time_point) %>%
  summarize(expression = median(expression)) %>%
  pivot_wider(
    id_cols = gene_id,
    names_from = time_point,
    values_from = expression
  ) %>%
  as_matrix
```

Next, we will use the `Mclust` package to estimate the number of clusters to split the dataset into. The package tests for a range of models and `k` values (number of clusters). We will test partitioning the dataset into up to 200 clusters, ie. `k = 1:200`. For each couple (model, cluster), calculates the Bayesian Information Content (BIC) value. It then picks the couple that maximizes the BIC score.

```{r calculate_k, eval=FALSE, cache=TRUE}
k_max <- 200

d_clust <- Mclust(
  norm,
  G = 1:k_max,
  modelNames = mclust.options("emModelNames")
)

plot(d_clust, what = "BIC")
```

```{r plot_d_clust, echo=FALSE, fig.width=7, fig.height=7, cache=TRUE}
d_clust <- readRDS(file = "bic/rds/k_max-200.rds")

plot(d_clust, what = "BIC")
abline(v = 21, lwd = 2)
abline(h = d_clust$BIC[21, "VVE"], lwd = 2)
points(
  21,
  d_clust$BIC[21, "VVE"],
  cex = 2,
  pch = 20
)
```

The best performing model is `VVE,21` (black dot in the plot above). This is a model using a multivariate Gaussian mixture model that uses ellipsoid kernels with equal orientations. We will proceed using the results from this clustering.

We will split our dataset in 21 clusters. On top of the genes we used to generate clusters, we will predict cluster assignment for all remaining genes.

```{r do_clustering, message=FALSE, fig.height=8.5, fig.width=15, cache=TRUE}

ct <- d_clust$classification
ct <- data.frame(gene_id = names(ct), cluster = ct)

newdata <- norm[!rownames(norm) %in% ct$gene_id,]

res <- predict(d_clust, newdata)

ct <- rbind(ct,  
            data.frame(
              gene_id = rownames(newdata),
              cluster = res$classification))

dat <- norm %>%
  as_tibble(rownames = "gene_id") %>%
  inner_join(ct) %>%
  pivot_longer(
    cols = -c(gene_id, cluster),
    names_to = "sample",
    values_to = "expression") %>%
  dplyr::rename(time_point = sample)

sds <- dat %>%
  group_by(time_point, cluster) %>%
  summarize(sd = sd(expression),
            expression = median(expression)) %>%
  mutate(gene_id = "none",
         sdmin = expression - sd,
         sdmax = expression + sd)

cluster_labels <- dat %>%
  group_by(cluster) %>%
  summarize(
    n = length(unique(gene_id))) %>%
  mutate(label = sprintf("cluster %d - n = %d", cluster, n)) %>%
  pull(label, name = cluster)

dat %>%
  mutate(time_point = factor(time_point, levels = colorder)) %>%
  ggplot(aes(x = time_point, group = gene_id)) +
  geom_line(mapping = 
              aes(y = expression), 
            linewidth = 0.25, 
            alpha = 0.33, 
            color = "grey"
  ) +
  geom_ribbon(
    data = sds,
    mapping = aes(x = time_point, ymin = sdmin, ymax = sdmax),
    fill = "dodgerblue",
    alpha = 0.66
  ) +
  geom_line(
    data = sds, 
    mapping = aes(x = time_point, y = expression), 
    color = "black") +
  facet_wrap(~ cluster, labeller = labeller(cluster = cluster_labels)) +
  theme_bw(base_size = 14) + 
  theme(aspect.ratio = 9/16,
        
        panel.grid = element_blank(),
        strip.background = element_blank())

```

We will proceed to characterize these clusters with GO enrichment.

# Gene Ontology Enrichment Analysis

## Clusters from time course analysis

```{r GO_enrichment_time_course, cache = TRUE, fig.height=20, fig.width=10, cache=TRUE}
time_based_clusters <- ct %>%
  as_tibble() %>%
  dplyr::mutate(gene_id = paste0("MGI:", gene_id))

universe <- all_deg %>%
  pull(gene_id) %>%
  unique() %>%
  paste0("MGI:", .)

ck_time_course_clustering <- compareCluster(gene_id ~ cluster, 
                     data = time_based_clusters, 
                     fun = enrichGO,
                     OrgDb = 'org.Mm.eg.db',
                     keyType = "MGI",
                     ont = "BP",
                     universe = universe,
                     minGSSize = 15)

ck_time_course_clustering <- setReadable(ck_time_course_clustering, OrgDb = org.Mm.eg.db, keyType = "MGI")

dotplot(ck_time_course_clustering, 
        x = "cluster", 
        showCategory = 2) + 
  scale_x_discrete(limits = c("2", "3", "4", 
                              "5", "6", "7", 
                              "8", "9", "11", 
                              "12", "14", "16", 
                              "18", "20", "21")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size=16),
        axis.title = element_text(size=16),
        axis.text.y = element_text(size=14))
```

Tables below report enriched GO terms for each cluster separately.

```{r split_ck_time_clusters, echo=FALSE, cache=TRUE}
splitted_time_based_clustering <- as.data.frame(ck_time_course_clustering) %>%
  group_by(Cluster) %>%
  group_split()
```

###  {.tabset}

#### Cluster 2

```{r go_cluster2, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[1]])
```

#### Cluster 3

```{r go_cluster3, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[2]])
```

#### Cluster 4

```{r go_cluster4, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[3]])
```

#### Cluster 5

```{r go_cluster5, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[4]])
```

#### Cluster 6

```{r go_cluster6, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[5]])
```

#### Cluster 7

```{r go_cluster7, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[6]])
```

#### Cluster 8

```{r go_cluster8, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[7]])
```

#### Cluster 9

```{r go_cluster9, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[8]])
```

#### Cluster 11

```{r go_cluster11, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[9]])
```

#### Cluster 12

```{r go_cluster13, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[10]])
```

#### Cluster 14

```{r go_cluster14, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[11]])
```

#### Cluster 16

```{r go_cluster16, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[12]])
```

#### Cluster 17

```{r go_cluster17, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[13]])
```

#### Cluster 19

```{r go_cluster19, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[14]])
```

#### Cluster 20

```{r go_cluster20, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[15]])
```

#### Cluster 21

```{r go_cluster21, echo=FALSE, cache=TRUE}
create_go_dt(splitted_time_based_clustering[[16]])
```

###  {.unnumbered}

## Pairwise comparisons

The visualization below shows the top 5 enriched GO term per comparison. Dots size corresponds to the ratio of genes found in cluster to be annotated with a given term, over the total number of genes annotated with that term. Dots color maps to significance level (p-value).

```{r clusterProfiler_compare_cluster, cache=TRUE, fig.width=10, fig.height=13}

clusters <- all_deg %>%
  dplyr::filter(deg) %>%
  dplyr::mutate(gene_id = paste0("MGI:", gene_id)) %>%
  dplyr::ungroup()

universe <- all_deg %>%
  pull(gene_id) %>%
  unique() %>%
  paste0("MGI:", .)

ck_pairwise_comparisons <- compareCluster(gene_id ~ time_point + direction, 
                     data = clusters, 
                     fun = enrichGO,
                     OrgDb = 'org.Mm.eg.db',
                     keyType = "MGI",
                     ont = "BP",
                     universe = universe,
                     minGSSize = 15)

ck_pairwise_comparisons <- setReadable(ck_pairwise_comparisons, OrgDb = org.Mm.eg.db, keyType = "MGI")

dotplot(ck_pairwise_comparisons, 
        x = "time_point",
        showCategory = 2) + 
  facet_grid(~direction, 
             labeller = labeller(
               direction = c(`1` = "upregulated",
                             `-1` = "downregulated"
                             ))) +
  scale_x_discrete(limits = c("0_30", "0_2", "0_4", 
                              "0_8", "0_24", "0_48", 
                              "0_7D", "0_WT", "WT_WT7D")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 16),
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14))
```

Tables below report enriched GO terms for each comparison, split by upregulated and downregulated.

```{r split_ck_pairwise, echo=FALSE, cache=TRUE}
splitted_pairwise_comparisons <- as.data.frame(ck_pairwise_comparisons) %>%
  group_by(Cluster) %>%
  group_split()
```

###  {.tabset}

#### 30 min vs 0 h

##### Upregulated

```{r 0_30_up, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[2]])
```

------------------------------------------------------------------------

##### Downregulated

```{r 0_30_down, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[1]])
```

#### 2 h vs 0 h

##### Upregulated

```{r 0_2_up, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[4]])
```

------------------------------------------------------------------------

##### Downregulated

```{r 0_2_down, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[3]])
```

#### 4 h vs 0 h

##### Upregulated

```{r 0_4_up, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[6]])
```

------------------------------------------------------------------------

##### Downregulated

```{r 0_4_down, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[5]])
```

#### 8 h vs 0 h

##### Upregulated

```{r 0_8_up, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[8]])
```

------------------------------------------------------------------------

##### Downregulated

```{r 0_8_down, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[7]])
```

#### 24 h vs 0 h

##### Upregulated

```{r 0_24_up, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[10]])
```

------------------------------------------------------------------------

##### Downregulated

```{r 0_24_down, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[9]])
```

#### 48 h vs 0 h

##### Upregulated

```{r 0_48_up, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[12]])
```

------------------------------------------------------------------------

##### Downregulated

```{r 0_48_down, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[11]])
```

#### 7 days vs 0 h

##### Upregulated

```{r 0_7d_up, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[14]])
```

------------------------------------------------------------------------

##### Downregulated

```{r 0_7d_down, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[13]])
```

#### 0 h vs WT 0 h

##### Upregulated

```{r 0_WT_up, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[16]])
```

------------------------------------------------------------------------

##### Downregulated

```{r 0_WT_down, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[15]])
```

#### WT 0 h vs WT 7 days

##### Upregulated

```{r WT_WT7d_up, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[18]])
```

------------------------------------------------------------------------

##### Downregulated

```{r WT_WT7d_down, echo=FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons[[17]])
```

###  {.unnumbered}

### Remove `WT_WT7D` DEG

We observe that some GO terms are enriched both in contrasts generated from genetically engineered samples activated with the drug (`0_*`) and WT treated with the activating drug (`WT_WT7D`).

We will filter out differentially expressed genes from the control case and calculate new GO enrichments.

```{r clusterProfiler_compare_cluster_filter_wt_deg, fig.width=10, fig.height=13, cache=TRUE}

fp <- all_deg %>%
  dplyr::filter(deg 
                & time_point == "WT_WT7D") %>%
  dplyr::pull("gene_id")

clusters <- all_deg %>%
  dplyr::filter(deg
                & !gene_id %in% fp) %>%
  dplyr::mutate(gene_id = paste0("MGI:", gene_id)) %>%
  dplyr::ungroup()

ck_pairwise_comparisons_no_wt_wt7d <- compareCluster(gene_id ~ time_point + direction, 
                     data = clusters, 
                     fun = enrichGO,
                     OrgDb = 'org.Mm.eg.db',
                     keyType = "MGI",
                     ont = "BP",
                     universe = universe,
                     minGSSize = 15)

ck_pairwise_comparisons_no_wt_wt7d <- setReadable(ck_pairwise_comparisons_no_wt_wt7d, OrgDb = org.Mm.eg.db, keyType = "MGI")

dotplot(ck_pairwise_comparisons_no_wt_wt7d, x = "time_point", showCategory = 2) + 
  facet_grid(~direction, 
             labeller = labeller(
               direction = c(`1` = "upregulated",
                             `-1` = "downregulated"
                             ))) +
  scale_x_discrete(limits = c("0_30", "0_2", "0_4", 
                              "0_8", "0_24", "0_48", 
                              "0_7D", "0_WT", "WT_WT7D")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size=16),
        axis.title = element_text(size=16),
        axis.text.y = element_text(size=14))
```

```{r split_ck_pairwise_filter_WT_WT7D, echo=FALSE, cache=TRUE}
splitted_pairwise_comparisons_no_WT_WT7D <- as.data.frame(ck_pairwise_comparisons_no_wt_wt7d) %>%
  group_by(Cluster) %>%
  group_split()
```

###  {.tabset}

#### 30 min vs 0 h

##### Upregulated

```{r go_WT_WT7D_filter_0_30_up, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[1]])
```

------------------------------------------------------------------------

##### Downregulated

```{r go_WT_WT7D_filter_0_30_down, echo = FALSE, cache=TRUE}
warning("No enriched terms.")
```

#### 2 h vs 0 h

##### Upregulated

```{r go_WT_WT7D_filter_0_2_up, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[3]])
```

------------------------------------------------------------------------

##### Downregulated

```{r go_WT_WT7D_filter_0_2_down, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[2]])
```

#### 4 h vs 0 h

##### Upregulated

```{r go_WT_WT7D_filter_0_4_up, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[5]])
```

------------------------------------------------------------------------

##### Downregulated

```{r go_WT_WT7D_filter_0_4_down, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[4]])
```

#### 8 h vs 0 h

##### Upregulated

```{r go_WT_WT7D_filter_0_8_up, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[7]])
```

------------------------------------------------------------------------

##### Downregulated

```{r go_WT_WT7D_filter_0_8_down, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[6]])
```

#### 24 h vs 0 h

##### Upregulated

```{r go_WT_WT7D_filter_0_24_up, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[9]])
```

------------------------------------------------------------------------

##### Downregulated

```{r go_WT_WT7D_filter_0_24_down, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[8]])
```

#### 48 h vs 0 h

##### Upregulated

```{r go_WT_WT7D_filter_0_48_up, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[11]])
```

------------------------------------------------------------------------

##### Downregulated

```{r go_WT_WT7D_filter_0_48_down, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[10]])
```

#### 7 days vs 0 h

##### Upregulated

```{r go_WT_WT7D_filter_0_7D_up, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[13]])
```

------------------------------------------------------------------------

##### Downregulated

```{r go_WT_WT7D_filter_0_7D_down, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[12]])
```

#### 0 h vs WT 0 h

##### Upregulated

```{r go_WT_WT7D_filter_0_WT_up, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[15]])
```

------------------------------------------------------------------------

##### Downregulated

```{r go_WT_WT7D_filter_0_WT_down, echo = FALSE, cache=TRUE}
create_go_dt(splitted_pairwise_comparisons_no_WT_WT7D[[14]])
```

###  {.unnumbered}

## Transcription factor activity inference 

```{r run_viper}

net <- get_collectri(organism = "mouse", split_complexes = TRUE)

mapping <- all_deseq2_results_tables %>% dplyr::select(gene_name, gene_id) %>% distinct()
deg_mat_gene_names <- deg_mat
rownames(deg_mat_gene_names) <- mapping$gene_name[match(rownames(deg_mat_gene_names), mapping$gene_id)]

sample_acts <- run_viper(mat = deg_mat_gene_names, 
                         network = net, 
                         .source = "source", 
                         .target = "target", 
                         .mor = "mor", 
                         minsize = 5)

```

```{r visualize_viper}

n_tfs <- 10

# Transform to wide matrix
sample_acts_mat <- sample_acts %>%
  pivot_wider(id_cols = 'condition', names_from = 'source',
              values_from = 'score') %>%
  column_to_rownames('condition') %>%
  as.matrix()

# Get top tfs with more variable means across clusters
tfs <- sample_acts %>%
  group_by(source) %>%
  summarise(std = sd(score)) %>%
  arrange(-abs(std)) %>%
  head(n_tfs) %>%
  pull(source)
sample_acts_mat <- sample_acts_mat[,tfs]

# Scale per sample
sample_acts_mat <- scale(sample_acts_mat)

# Choose color palette
palette_length <- 100
my_color <- colorRampPalette(c("darkblue", "white","red"))(palette_length)

my_breaks <- c(seq(-3, 0, length.out = ceiling(palette_length/2) + 1),
               seq(0.05, 3, length.out = floor(palette_length/2)))

colorder <- paste0(rep(c("0h", "30min", "2h", "4h", "8h", "24h", "48h", "7D"), 
                       each = 3), 
                   "_rep", 
                   1:3)

# Plot
pheatmap(sample_acts_mat[colorder, ], border_color = NA, color = my_color, breaks = my_breaks, 
         cluster_rows = FALSE) 

```


# Transposable Elements

We will import gene expression quantification as computed using the STARTE-random methodology (@Teissandier2019). In brief, we perform standard trimming with Trimmomatic, mapping with STAR using a set of parameters permissive for multimapping reads, and FeatureCounts to quantify expression of each TE family.

```{r import_te_data, cache=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
get_results_table <- function(file_path){
  tp <- basename(dirname(dirname(file_path)))
  
  read_csv(file_path) %>%
    dplyr::select(-1) %>%
    dplyr::mutate(time_point = tp)
}

mapping_table <- read_delim("3t-seq/data/references/rmsk/mm10.gtf", 
                            col_names = FALSE) %>%
  dplyr::select(X9) %>%
  dplyr::mutate(
    repName = sub(
      "^repName=(.+);repClass=.+;repFamily=.+;repStart=.+;repEnd=.+;repLeft=.+$", 
      "\\1", 
      X9
    ),
    repClass = sub(
      "^repName=.+;repClass=(.+);repFamily=.+;repStart=.+;repEnd=.+;repLeft=.+$", 
      "\\1", 
      X9
    ),
    repFamily = sub(
      "^repName=.+;repClass=.+;repFamily=(.+);repStart=.+;repEnd=.+;repLeft=.+$", 
      "\\1", 
      X9
    )
  ) %>%
  dplyr::select(-X9) %>%
  dplyr::distinct()

lfc_table <- list.files("3t-seq/data/results/alignments/starTE",
           pattern = "lfc.txt",
           full.names = TRUE,
           recursive = TRUE) %>%
  purrr::map(get_results_table) %>%
  purrr::reduce(bind_rows) %>%
  dplyr::left_join(
    mapping_table, 
    by = c("gene_name" = "repName"), 
    multiple="all"
    )

```

## TE from @Dura2022

We will pick TE as shown in [Figure 1F](https://www.nature.com/articles/s41588-022-01040-z/figures/1) of @Dura2022.

Unfortunately, not all displayed TE could be detected in this dataset. The list below shows the missing genes.

```{r select_te, cache=TRUE}

line1 <- c("L1MdA_I", "L1MdA_II", "L1MdA_III", "L1MdA_IV", 
           "L1MdA_V", "L1MdA_VI", "L1MdA_VII", "L1MdTf_I",
           "L1MdTf_II", "L1MdTf_III", "L1MdF_II", "L1MdGf_I", 
           "L1MdGf_II")

sine <- c("B2_Mm1a", "B1_Mm")

erv1 <- c("MURVY-int", "MMERGLN-int", "MMERGLN_LTR")

ervk <- c("MMERVK10C-int", "IAPEz_int", "IAPEy-int", "IAPd-int",
          "ETnERV2-int", "ETnERV3-int", "RLTR10-int")

ervl <- c("ORR1A1-int", "ORR1A0-int", "ORR1A0", "MERVL-int")

selected <- c(line1, sine, erv1, ervk, ervl)

setdiff(selected, lfc_table$gene_name)
```

We will first produce a heatmap similar to the Figure from the paper above.

```{r plot_te_heatmap, fig.width=10, fig.height=5, cache=TRUE}
as_matrix <- function(df) {
  rn <- df$gene_name
  mat <- as.matrix(df[,-1])
  rownames(mat) <- rn
  return(mat)
}

mat <- lfc_table %>%
  dplyr::filter(gene_name %in% selected) %>%
  pivot_wider(names_from = time_point,
              values_from = log2FoldChange,
              id_cols = gene_name) %>%
  as_matrix()

available_te <- intersect(selected, rownames(mat))

mat <- mat[available_te, c("0_30", "0_2", "0_4", 
                           "0_8", "0_24", "0_48", 
                           "0_7D", "0_WT", "WT_WT7D")]

rowann <- data.frame(
  row.names = available_te,
  class = mapping_table %>% 
    dplyr::filter(repName %in% available_te) %>% 
    dplyr::pull("repClass"),
  family = mapping_table %>% 
    dplyr::filter(repName %in% available_te) %>% 
    dplyr::pull("repFamily")
)

pheatmap(t(mat),
         color = colorRampPalette(
           rev(brewer.pal(n = 5, name = "RdBu")))(100),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         annotation_col = rowann)
```

We will also produce a line plot showing the same data.

```{r plot_te_lines, fig.height=10, fig.width=10, warning=FALSE, message=FALSE, cache=TRUE}
lfc_table %>%
  dplyr::filter(gene_name %in% selected) %>%
  ggplot2::ggplot(aes(time_point, log2FoldChange, group = gene_name)) +
  geom_line() +
  facet_wrap( ~ repClass + repFamily) +
  theme_bw(base_size = 16) +
  theme(aspect.ratio = 1, 
        strip.background = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## All differentially expressed TE

We will produce similar visualizations using differentially expressed TE from DESeq2 analysis as performed by the 3t-seq pipeline.

Fist we will call DE TE using thresholds defined above.

```{r call_te_deg, cache=TRUE}
te_de <- lfc_table %>%
  dplyr::mutate(deg = log2FoldChange > lfc_thr
               & ifelse(is.na(padj), pvalue < pval_thr, padj < pval_thr))
```

We will produce a clustered heatmap.

```{r pheatmat_te_de, fig.height=10, fig.width=7, cache=TRUE}

mat <- te_de  %>%
  dplyr::filter(deg & gene_name != "Eulor5A") %>%
  tidyr::pivot_wider(
    id_cols = gene_name, 
    values_from = log2FoldChange, 
    names_from = time_point, 
    values_fill = NA) %>%
  as_matrix 

rowann <- data.frame(
  row.names = rownames(mat),
  class = mapping_table %>% 
    dplyr::filter(repName %in% rownames(mat)) %>% 
    dplyr::pull("repClass"),
  family = mapping_table %>% 
    dplyr::filter(repName %in% rownames(mat)) %>% 
    dplyr::pull("repFamily")
)

mat <- mat[, c("0_30", "0_2", "0_4", 
               "0_8", "0_24", "0_48", 
               "0_7D", "0_WT", "WT_WT7D")]

pheatmap(mat,
         color = colorRampPalette(
           brewer.pal(n = 5, name = "Purples"))(100),
         na_col = "grey70", 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         annotation_row = rowann,
         show_rownames = FALSE)
```

It looks like pairwise calls for differentially expressed TE result in sparse data. Temporal clustering will do better?

# References
